---
title: "Survival analysis of Games of Thrones characters"
output:
  html_document:
    df_print: paged
---
Author: Mariano Crimi
Date: 26-Feb-2020
Course: Survival Analysis
Methodology: SPOC

#Introduction


```{r}
library(readr)
library(tidyverse)
library(survival)
library(survivalROC)
library(glmnet)
library(Matrix)
library(ggplot2)
library(survminer)
library(gridExtra)
library(RColorBrewer)
palette <- brewer.pal(3, "Pastel2") 
```

**Read in the dataset file**

```{r}
datRaw<-read.csv("character_data_S01-S08.csv")
```

**Data wrangling **
Drop variables that are of no interest and convert categorical variables to factors.

```{r}
drops <- c("name",
           "dth_season",
           "dth_episode", 
           "dth_time_sec", 
           "dth_time_hrs",
           "dth_description",
           "intro_season",
           "intro_episode",
           "intro_time_hrs",
           "intro_time_sec",
           "featured_episode_count",
           "exp_episode",
           "exp_time_hrs",
           "exp_season",
           "censor_time_sec",
           "censor_time_hrs",
           "icd10_dx_code",
           "icd10_dx_text",
           "icd10_cause_code",
           "icd10_cause_text",
           "icd10_place_code",
           "icd10_place_text",
           "top_location",
           "geo_location",
           "time_of_day",
           "X",
           "X.1",
           "X.2",
           "X.3",
           "X.4",
           "X.5")
dat <- datRaw[ , !(names(datRaw) %in% drops)]

dat <- mutate(dat,
              sex = factor(sex, levels = c(1,2), 
                                     labels = c("Male", 
                                                "Female")),
              social_status = factor(social_status, levels = c(1,2), 
                                                    labels = c("Highborn", 
                                                               "Lowborn")),
              occupation = factor(occupation, levels = c(1,2,9), 
                                              labels = c("Silk collar", 
                                                         "Boiled leather collar",
                                                         "Unknown")),
              allegiance_switched = factor(allegiance_switched, levels = c(1,2), 
                                                                labels = c("No", 
                                                                            "Yes")),              
              religion = factor(religion, levels = c(1,2,3,4,5,6,7,9 ), 
                                          labels = c("Stallion",
                                                     "LoL",
                                                     "Faithof7",
                                                     "OldGods",
                                                     "D.God",
                                                     "M.FacedGod",
                                                     "Other",
                                                     "Unknown")),
              allegiance_last = factor(allegiance_last, levels = c(1:9), 
                                                        labels = c("Stark",
                                                                   "Targaryen",
                                                                    "NightWatch",
                                                                    "Lannister",
                                                                    "Greyjoy",
                                                                    "Bolton",
                                                                    "Frey",
                                                                    "Other",
                                                                    "Unclear")))
```

#Exploration of the data

```{r}

grid.arrange(qplot(dat$allegiance_last, xlab = "Allegiance"),
             qplot(dat$sex, xlab = "Sex"),
             qplot(dat$social_status, xlab = "Social Status"),
             qplot(dat$occupation, xlab = "Ocuppation"),
             qplot(dat$allegiance_switched, xlab = "Switched allegiance?"),
             qplot(dat[dat$religion != "Unknown",]$religion, xlab = "Religions (removing unknowns)"),
             qplot(dat$exp_time_sec/60, xlab = "Survival in mins", ylab="y"),
             qplot(factor(dat$dth_flag, levels = c(0:1), labels = c("Yes","No")), xlab="Death"),
             qplot(dat$prominence, xlab = "Prominence"),
             nrow = 5
            )
summary(dat)
dim(dat)
```

**Prominence:** Given that the metric was calculated as a function of the survial of the episodes (: [prominence] = ([featured_episode_count] / [exp_episode]) * [exp_season]) I was initially expecting that this variable would add very little information to the model.
Upon checking the correlation level seems acceptable to fit it into the model

```{r}
cor(dat$prominence,dat$exp_time_sec/60)
qplot(x=(dat$prominence),y=(dat$exp_time_sec/60), geom = c("jitter","smooth"), xlab="Prominence score" , ylab="Survival Time  (mins)")
```

I now proceed to add prominance as a three level factor, high-medium-low
```{r}
dat$prominence<- cut(dat$prominence, breaks=c(0,1,4,9), label=c("low","med","high"))
qplot(dat$prominence, xlab = "Prominence count")
```

We then create a survival object to be used as the response vairable of our dataset using survival::Surv.
Object Y will consist on Time (right censored) *in minutes* and Event (0=alive, 1=dead)

```{r}
dat$y <- with(dat, Surv(time= exp_time_sec/60, event=dth_flag))
```


#Survival Analysis

We choose to use non-parametric modelization of the survival. In particular we chose Kaplen Mayer estimator (Nelson Allen gives us very similar results). 
We first explore the general survival curve of the dataset. We get an estimated median overall survival of 1729 minutes (with 95% confidence that the actual median is within 1215 mins and 2546 mins)

```{r}
fitKM<-survfit(y ~ 1, data = dat)
ggsurvplot(fitKM, xlab="Time(mins)", ylab="Survival Probability",main="KM Plot", surv.median.lin="hv")
fitKM
summary(fitKM)

```

We then proceed to explore all the covariates individually

```{r}

fitSex <- survfit(y ~ sex, data = dat)
fitOccupation <- survfit(y ~ occupation, data = dat)
fitAllegiance <- survfit(y ~ allegiance_last, data = dat)
fitSwitch <- survfit(y ~ allegiance_switched, data = dat)
fitReligion <- survfit(y ~ religion, data = dat)
fitProminence <- survfit(y ~ prominence, data = dat)
fitStatus <- survfit(y ~ social_status, data = dat)



ggsurvplot(fitSex, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitOccupation, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitAllegiance, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitSwitch, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitReligion, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitStatus, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)
ggsurvplot(fitProminence, xlab="Time(mins)", ylab="Survival Probability", pval=TRUE, pval.method = TRUE)


```

At this point we would drop intro_season, occupation and social status as good predictors.
We now explore the relationship between the most interesting factors

For example we examine religion which at a first glance seems siginificant (p=0.002)
We llok at the proportion of other interesting factors within the groups
```{r}
#plot(table(dat$religion,dat$sex), main="Proportion of sex per religion")
#plot(table(dat$religion,dat$allegiance_last), col=105:110, main="Proportion of allegiance per religion")
ggplot(dat, aes(religion, fill=sex)) + geom_bar()
ggplot(dat, aes(religion, fill=allegiance_last)) + geom_bar()


```
We see that females and some allegiances are largely underrepresented in some religions. Let us stratify by sex and see if we find the same significance

```{r}
survdiff(y ~ religion+strata(sex,allegiance_last), data=dat)
```


Occupation analysis
```{r}
table(dat$occupation,dat$sex)
plot(table(dat$occupation,dat$sex), main="Proportion of sex per occupation", col=palette)
survdiff(y ~ occupation+strata(sex), data=dat)
survdiff(y ~ occupation, data=dat)

```

Social status analysis
```{r}

survdiff(y ~ social_status, data=dat)

plot(table(dat$allegiance_last,dat$social_status),   main="Proportion of social status per allegiance", col=palette)
coxph(y ~ allegiance_last, data = dat)

plot(table(dat$prominence,dat$social_status),   main="Proportion of social status per prominence", col=palette)
coxph(y ~ prominence, data = dat)

survdiff(y ~ social_status+strata(prominence,allegiance_last), data=dat)


```


Prominence and allegiance_switched
```{r}
survdiff(y ~ prominence+strata(allegiance_switched), data=dat)
survdiff(y ~ allegiance_switched+strata(prominence), data=dat)

plot(table(dat$prominence,dat$allegiance_switched),   main="Proportion desertion per prominence", col=palette)


```
Others

```{r}
survdiff(y ~ allegiance_last+strata(sex,social_status,prominence,allegiance_switched), data=dat)
survdiff(y ~ sex+strata(allegiance_last,allegiance_switched,prominence,social_status), data=dat)
survdiff(y ~ allegiance_switched+strata(sex,prominence,allegiance_last,social_status), data=dat)
survdiff(y ~ prominence+strata(allegiance_last,sex,allegiance_switched,social_status), data=dat)
```


We become more and more convinced that these are good predictors:

*prominence
*sex
*allegiance_switched
*allegiance_last

Before proceeding with the manual model creation, we would like to perform a likelihood ratio test on the nested model
to test the assumption that the religon we removed due to stratification wouldn't add significance to the model
```{r}
fitCompNestA <- coxph(y ~ sex+allegiance_last+allegiance_switched+prominence, data = dat)
fitCompFullB <- coxph(y ~ sex+allegiance_last+allegiance_switched+prominence+occupation, data = dat)
anova(fitCompNestA, fitCompFullB)

fitCompNestC <- coxph(y ~ sex+allegiance_last+allegiance_switched+prominence, data = dat)
fitCompFullD <- coxph(y ~ sex+allegiance_last+allegiance_switched+prominence+social_status, data = dat)
anova(fitCompNestC, fitCompFullD)

```
In both cases we can't reject H0= social status/occupation is 0, therefore we assume these doesn't predictive information to the model


#Diagnostics
_____________

We now check the **proportionality of hazards** assumption by plotting the survival functin curves in
complementary loglog scale

```{r}
#Plot log-log curves
plot(survfit(y ~ prominence, data = dat), fun= "cloglog", col = 1:3, main="prominence risks (cloglog)")
plot(survfit(y ~ sex, data = dat), fun= "cloglog", col = 1:2,main="sex risks (cloglog)")
plot(survfit(y ~ allegiance_switched+strata(prominence), data = dat), fun= "cloglog", col = 1:6,main="allegiance_switched risks (cloglog)")
plot(survfit(y ~ allegiance_last, data = dat), fun= "cloglog", col = 1:9,main="allegiance_last risks (cloglog)")

#Hypothesis testing
cox.zph(coxph(y ~ sex+allegiance_switched+allegiance_last+prominence, data = dat))
#Plot
plot(cox.zph(coxph(y ~ sex+allegiance_switched+allegiance_last+prominence, data = dat)))
ggcoxzph(cox.zph(coxph(y ~ sex+allegiance_switched+allegiance_last+prominence, data = dat)))
```

Stratification
```{r}
#Alteratives
cox.zph(coxph(y ~ sex+prominence+allegiance_last+strata(allegiance_switched), data = dat))
cox.zph(coxph(y ~ sex+prominence+allegiance_switched+strata(allegiance_last), data = dat))
cox.zph(coxph(y ~ prominence+allegiance_switched+allegiance_last+strata(sex), data = dat))
#The winner
cox.zph(coxph(y ~ sex+allegiance_switched+allegiance_last+strata(prominence), data = dat))

```

```{r}
fitAnalysis<-coxph(y ~ sex+allegiance_switched+strata(prominence), data=dat)
```


```{r}
#Martingale residuals
dat_res<-dat
dat_res$residualsM <- residuals(fitAnalysis, type = "martingale")
dat_res$residualsB <- residuals(fitAnalysis, type = "dfbetas")
with(dat_res, {
  
  boxplot(residualsM ~ sex)

  boxplot(residualsM ~ allegiance_switched)
  

})

```
Outliers analysis
```{r}
dat_res$residualsB <- sqrt(rowSums(dat_res$residualsB^2))
```

```{r}
plot(dat_res$residualsB, type = 'h', ylab="dfbetas")
```


We see that only sex passes the test for proportionality of hazards. Looking at the residuals it looks that some artificial censoring would help some of the variables
```{r}
new_censor_mins=300
new_censor_sec=new_censor_mins*60
dat_trunc <- within(dat, {
  dth_flag_truncated <- ifelse(exp_time_sec > new_censor_sec, 0, dth_flag)
  exp_time_sec_truncated <- ifelse(exp_time_sec > new_censor_sec,new_censor_sec, exp_time_sec)
})
  
cox.zph(coxph(Surv(exp_time_sec_truncated/60, dth_flag_truncated) ~ sex+prominence+allegiance_switched+allegiance_last, data = dat_trunc))

```

#Conclusion
```{r}
summary(fitAnalysis)
```

#Regression

Data preparation for regression
```{r}
set.seed(123) # setting seed for reproducibility

#Leave only explanatory varialbes
dat_models <- dat[,!(names(dat) %in% c("id","dth_flag","exp_time_sec"))]

#Training and testing sets
trn_size <- floor(0.75 * nrow(dat_models))
train_ind <- sample(seq_len(nrow(dat_models)), size = trn_size)
dat_trn <- dat_models[train_ind, ]
dat_test <- dat_models[-train_ind, ]

```

#Manual model
Based on the invividual log rank tests, the anova tests and the proportinality of hazards we leans towards a model that considers sex and allegiance_switched, stratifying for prominence.
```{r}
fitManual <- coxph(y ~ sex+allegiance_switched+strata(prominence), data=dat_trn)
summary(fitManual)
betasManual <- coef(fitManual)
```

#Feature Selection
Data preparation
```{r}
y <- dat_trn$y
Xmatrix <- model.matrix(y ~ ., data = dat_trn)
Xmatrix <-Xmatrix[,-c(1)] # remove intercept
```

#ElasticNet
## Using 1se and min cv error lambdas

```{r}
fitElastic1se <- cv.glmnet(x=Xmatrix, y=y, family = "cox")
betasElasticNet1se.all <- coef(fitElastic1se, s = "lambda.1se")
betasElasticNet1se <- betasElasticNet1se.all[betasElasticNet1se.all != 0]
names(betasElasticNet1se) <- colnames(Xmatrix)[as.logical(betasElasticNet1se.all != 0)]
names(betasElasticNet1se)

#Plot lambdas
plot(fitElastic1se)


fitElasticMin <- cv.glmnet(x=Xmatrix, y=y, family = "cox")
betasElasticNetMin.all <- coef(fitElasticMin, s = "lambda.min")
betasElasticNetMin <- betasElasticNetMin.all[betasElasticNetMin.all != 0]
names(betasElasticNetMin) <- colnames(Xmatrix)[as.logical(betasElasticNetMin.all != 0)]
names(betasElasticNetMin)

```

#Stepwise using AIC

```{r}
fitAic<- step(coxph(y ~ ., data = dat_trn),direction = "both")
betasAIC <- coef(fitAic)
summary(fitAic)
```


#Model Selection and performance evaluation

```{r}
#Function for linear combinantion of the coefficients
lincom <- function(b, Xr) rowSums(sweep(Xr[, names(b), drop = FALSE], 2, b, FUN = "*"))



models_coefficients <- tibble(
  method = c("manual", "aic", "elasticNet1se","elasticNetMin"),
  coefficients = list(betasManual, betasAIC, betasElasticNet1se,betasElasticNetMin)
)



dat_testMatrix <- model.matrix(y ~ .-1, data=dat_test)
ytest <- dat_test$y


got_models_performance <- mutate(models_coefficients,
                             predictions = map(coefficients, ~ lincom(., dat_testMatrix)),
                             cox_obj = map(predictions, ~ coxph(ytest ~ I(. / sd(.)))),
                             cox_tab = map(cox_obj, broom::tidy)
) %>% unnest(cox_tab)


#Test against actual values
got_models_performance <- mutate(got_models_performance,
                             AUC = map_dbl(predictions, ~ survivalROC::survivalROC(Stime=ytest[, 1], status=ytest[, 2], ., predict.time = 3000, method = "KM")$AUC)
)
got_models_performance
got_models_performance_html<- mutate(got_models_performance,
                                     estimate = round(estimate,4),
                                     statistic = round(statistic,4),
                                     p.value = round(p.value,4),
                                     conf.low = round(conf.low	,4),
                                     conf.high = round(conf.high,4),
                                     AUC = round(AUC,4)
                                     )

got_models_performance %>%select(method, estimate, std.error, p.value, AUC)
got_models_performance$cox_obj[4]
betasElasticNetMin
```

#ROC comparison (Manual vs ElasticNetMin)

```{r}
ROC_Manual<-survivalROC(Stime=ytest[, 1], status=ytest[, 2], marker= unlist(got_models_performance$predictions[1], use.names=FALSE), predict.time = 3000, method = "KM")

ROC_Elastic_min<-survivalROC(Stime=ytest[, 1], status=ytest[, 2], marker= unlist(got_models_performance$predictions[4], use.names=FALSE), predict.time = 3000, method = "KM")

ROC <- list(Manual = ROC_Manual, elasticNetMin = ROC_Elastic_min)
map_dbl(ROC, "AUC")

dfl <- map(ROC, ~ with(., tibble(cutoff = cut.values, FP, TP)))
for(nm in names(dfl)) {
  dfl[[ nm ]]$marker <- nm
}
ROCdat <- do.call(rbind, dfl)

ggplot(ROCdat, aes(FP, TP, color = marker)) +
  geom_line() +
  theme_bw(base_size = 9)
```






